\bfseries\LARGE\singlespacing{A novel spatiotemporal model to estimate \emph{Mycobacterium tuberculosis} transmission risks in crowded indoor settings using environmental, clinical, and patient movement data}

### 2024-05-16 doc/paper/manuscript.tex

\title{\bfseries\LARGE\singlespacing{A novel spatiotemporal model to estimate \emph{Mycobacterium} tuberculosis transmission risks in crowded indoor settings using environmental, clinical, and patient movement data}}
% author list
\author[1]{Nicolas Banholzer}
\author[2]{Keren Middelkoop}
\author[2]{Juane Leukes}
\author[3]{Ernest Weingartner}
\author[1]{Remo Schmutz}
\author[1]{Kathrin Zürcher}
\author[1]{Matthias Egger}
\author[2]{Robin Wood}
\author[1*]{Lukas Fenner}

\affil[1]{Institute of Social and Preventive Medicine, University of Bern, Bern, Switzerland}
\affil[2]{Desmond Tutu HIV Centre, Department of Medicine, University of Cape Town, Cape Town, South Africa}
\affil[3]{Institute for Sensors and Electronics, University of Applied Sciences and Arts Northwestern Switzerland, Windisch, Switzerland}

@@

@@

% \noindent\textbf{S1 Appendix:} Includes supplementary text, tables and figures.

@@

%TC:newcounter abst Words in abstract
%TC:envir abstract [] abst
\begin{abstract}\normalfont
\noindent\textbf{Background:} Tuberculosis (TB) caused by \emph{Mycobacterium tuberculosis (Mtb)} is strictly airborne and thus the risk of transmission is higher in crowded indoor environments, especially healthcare facilities. Existing approaches to modeling the risk of airborne transmission typically assume that the airspace is well mixed, not considering that proximity to infectious individuals may increase the risk of transmission. We aimed to develop a  spatiotemporal model of airborne \emph{Mtb} transmission using environmental, clinical, and patient movement data to estimate TB transmission during the COVID-19 pandemic in healthcare facilities and to evaluate the impact of infection control measures.

\noindent\textbf{Methods:} We collected environmental (CO$_2$ levels), clinical (patient visits and TB disease status), and patient movement data (anonymous patient tracking from video sensors) in a primary care clinic in Cape Town, South Africa, for five days in October/November 2021 during the COVID-19 pandemic when universal use of masks was compulsory and ventilation was increased. We matched patient movements with clinical records to identify the spatiotemporal location of  infectious TB patients. We modeled the spatiotemporal concentration of infectious doses (quanta) and estimated the patient-specific risk of infection using the Wells-Riley equation.

\noindent\textbf{Results:} Clinical records registered 894 patients, four of whom were diagnosed with TB. Video sensors detected 1,438 clinic attendees with a median stay of 25\,min (interquartile range [IQR] 13\,min$-$46\,min). CO$_2$ levels were low (median 431\,ppm, IQR 406\,ppm$-$458\,ppm), compatible with high ventilation rates. The modeled risk of infection per clinic attendee was 0.059\% (95\%-credible interval [CrI] 0.001\%$-$0.402\%). The risk would have increased 1.28-fold (95\%-CrI 1.19$-$1.36) without mask use and 1.74-fold (95\%-CrI 0.47$-$5.34) with lower, pre-pandemic ventilation rates. Furthermore, the individual risk was significantly associated with the number of close-contact encounters and time spent in the clinic.

\noindent\textbf{Conclusions:} Our novel spatiotemporal transmission model estimated the risk of \emph{Mtb} transmission in a South African clinic as low during the COVID-19 pandemic, mainly due to infection control measures. Our spatiotemporal modeling could be used to identify high-risk areas and evaluate the impact of infection control measures in clinics.

@@

\noindent\textbf{Keywords:} Mycobacterium tuberculosis, airborne infection, patient tracking, spatiotemporal model
% maximum of 3-5 keywords
\newpage

@@

%TC:break main
\section{Introduction}

% TB and the route of transmission
Tuberculosis (TB) is one of the leading causes of death worldwide, especially in the South African region, which is disproportionately affected by the TB burden. After two years of COVID-19-related disruption in the progress made in TB prevention and control before the pandemic, the impact of the pandemic is about to reverse \cite{WHO2022TBReport}. However, global TB targets have 
been largely missed (REF).  \emph{Mycobacterium tuberculosis} (\emph{Mtb}), the causative agent of TB, is transmitted via respiratory particles in the exhaled air of infectious persons\cite{Rieder1999,Patterson2021Tuberculosis}. \emph{Mtb} is primarily carried in smaller particles in the size range of $1-7\mu$m\cite{Fennelly2020Lancet}, which can survive in the air for multiple hours\cite{Loudon1969AMRRD}. Survival of airborne pathogens is greater in crowded, poorly ventilated indoor environments\cite{Rieder1999,CPS2013Book,Nardell1991ARRD,Wang2021Science,Morawska2021}. As such, healthcare facilities present a particularly high-risk setting because both more infectious and more susceptible individuals are also more likely to visit a clinic\cite{McCreesh2020IJTLD}. A modeling study in South Africa estimated that 4\% to 14\% of adult TB cases originate from primary care clinics\cite{McCreesh2022BMJGlobalHealth}.

% the Wells-Riley model and its limitations
The Wells-Riley model\cite{Riley1978AJE} is widely used to estimate the risk of airborne transmission in a variety of indoor settings\cite{Andrews2014JID,Taylor2016IJTLD,Hella2017JInfect,Zemouri2020JDR}, including primary care clinics\cite{Zurcher2022JID,McCreesh2021BMJGlobalHealth}. The model estimates the risk of infection based on the number of infectious individuals in space, the generation rate of infectious quanta (doses of pathogen-carrying particles), the breathing rate per person, and the outdoor air supply rate. The Wells-Riley model, and variations thereof\cite{Rudnick2003IndoorAir}, assume a well-mixed airspace, which means that the quanta concentration is the same throughout the room. However, the quanta concentration is typically higher near the infectious source\cite{Wang2021Science,Vuorinen2020SafSci,Chen2020BuildEnv}, and previous studies have shown that the risk of TB infection is associated with proximity to these individuals\cite{Ko2004RiskAnal,Kenyon1996NEJM}.

% what this study adds
We developed a spatiotemporal extension of the Wells-Riley model to model the concentration of infectious quanta in space and time. We combined clinical and person tracking data to determine the spatiotemporal location of TB patients and susceptible attendees at a primary care clinic in South Africa for five days in October/November 2021. We also measured indoor CO$_2$ levels to model the spatial diffusion and removal of infectious quanta over time. We used Monte Carlo simulation to estimate the risk of infection per clinic attendee. Finally, we assessed the impact of infection control measures in place during the study in response to the COVID-19 pandemic (mask use and increased ventilation) by modeling the risk of infection under hypothetical pre-pandemic scenarios without universal mask use and pre-pandemic ventilation conditions in the clinic.

@@

\section{Methods}

\subsection{Study design}

We build on the design of a pilot study in 2019\cite{Zurcher2022JID}, which was described in detail in a study protocol \cite{Zurcher2020BMJ}. We collected environmental data (indoor CO$_2$ levels), clinical data (TB status), and video sensor data (patient movements) for five days during the COVID-19 pandemic in October/November 2021 (October 13, 15, 25, and November 4, 5) at a primary care clinic in Cape Town, South Africa.

\subsection{Study setting}

The primary care clinic offers both TB and HIV services as well as other basic clinical services, Monday through Friday, from 8\,am to 4\,pm. The clinic is located in a large settlement of formal and semi-formal housing where TB and HIV are highly prevalent\cite{Wood2007AMJRCCD,Middelkoop2011JAIDS}. We defined three areas within the clinic (\Cref{fig:floor-plan}): the waiting room (10.55m $\times$ 5.50m $\times$ 3.00m), corridor (12.45m $\times$ 2.20m $\times$ 2.50m), and TB room (4.75m $\times$ 3.50m $\times$ 3.00m). The TB room was mainly used for the treatment of diagnosed TB patients, but also for the screening of suspected patients with respiratory symptoms who were ultimately not diagnosed with TB.

\begin{figure}[!htpb]
    \centering
    \includegraphics{doc/clinic-schematic-annotated-view.pdf}
    \caption{\textbf{Schematic view of the clinic}. Patients typically enter the clinic through the entrance on the bottom left, register at the reception, then wait in the waiting room or along the corridor until they see a doctor in one of the care rooms or the TB room, and finally exit the clinic through one of the exits in the waiting room or at the end of the corridor. Colored icons show the placement of the video sensors to track patient movements and the devices monitoring CO$_2$ levels.}
    \label{fig:floor-plan}
\end{figure}

\subsection{Data}

\subsubsection{Environmental data}

Four devices monitored indoor CO$_2$ levels (Digital CO$_2$ Monitor Carbon Dioxide Meter XE-2000, XEAST, Guangdong, China) in the waiting room, corridor, and TB room (\Cref{fig:floor-plan}). CO$_2$ levels were recorded in parts per million (ppm) at 1min intervals. Missing values were linearly imputed. On October~13, the corridor's CO$_2$ levels were missing and imputed with the waiting room's CO$_2$ levels because their CO$_2$ measurements were similar on the other days.

\subsubsection{Clinical data}

We extracted clinical data from the electronic patient registry for all patients who attended the clinic during the study. Clinical records included the date and time of arrival, TB diagnostic results, and date of TB treatment initiation. We defined infectious patients as individuals who tested bacteriologically positive for TB during the study or who had received TB treatment for less than four weeks. Patients on treatment for more than four weeks or who had recently completed treatment were considered neither infectious nor susceptible. We also defined suspected TB patients as individuals who were tested for TB following symptom screening, but who were ultimately not diagnosed with TB.

\subsubsection{Person tracking data}

We used an anonymous personal tracking system using video sensors (Xovis, Zollikofen, Switzerland) to monitor the movements of people (clinic staff, patients, and other visitors) throughout the clinic at 1~second intervals (\Cref{fig:floor-plan}). The resulting time-stamped tracking data consisted of a person’s height, their position recorded as x-y coordinates, and a unique ID for each person's track while in the clinic. \supp~Figure~\zref{fig:tracking-examples} shows some sample tracks from our study. Note that people in the clinic could contribute multiple tracks if they moved out of the range of a sensor or if they were briefly unrecognized as a person. To address these issues, we created an R Shiny tool to manually reconnect interrupted tracks back together (\supp~Text~\zref{sec:setting-and-data}). We also labeled the tracks to identify movements of the clinic staff.

\subsection{The Wells-Riley equation}

Our spatiotemporal model is based on the Wells-Riley equation\cite{Riley1978AJE}, which estimates the risk (probability in \%) of infection $P$ as
\begin{align}
    P = \frac{D}{S} = 1 - \exp\left(\frac{Ipqt}{Q}\right),
\end{align}
where $D$ is the number of diseased cases, $S$ is the number of susceptible cases, $I$ is the number of infectious people in the indoor space, $p$ is the breathing rate (m$^3$ h$^{-1}$), $q$ is the quantum (infectious dose) generation rate (quanta h$^{-1}$), $t$ is the exposure time (h), and $Q$ is the ventilation rate (m$^3$ h$^{-1}$). The probability of infection is estimated with a Poisson relation, considering the stochastic behavior of airborne infection, where one quantum corresponds to a 67\% risk of infection. The main assumption of the Wells-Riley model is a well-mixed airspace, which means that the quanta concentration $N = Iq/Q$ (quanta m$^{-3}$) is the same everywhere in the room. Instead, our study considers a mixed airspace with spatial variation in the quanta concentration. This is accomplished using patient movement data combined with clinical records to identify the spatiotemporal location of infectious patients who generate quanta.

\subsection{Spatiotemporal modeling approach}

A detailed description of the Wells-Riley model and our spatiotemporal extension is provided in \supp~Text~\zref{sec:spattemp-model}. In the following, we describe the main components of our modeling approach, which are summarized in \Cref{fig:modeling-flow}.

\begin{figure}[!htpb]
    \centering
    \includegraphics{doc/model-flow-chart.pdf}
    \caption{\textbf{Visual summary of the spatiotemporal modeling approach}. (1)~Clinical and person tracking data were combined to determine the spatiotemporal location of diagnosed (infectious) TB patients. (2)~Both diagnosed and randomly selected undiagnosed  TB among clinic attendees generate infectious quanta. (3)~Tracking and environmental data were combined to estimate the air change rate. (4)~Air change rate was related to the speed at which quanta diffuses indoors. (5)~The air change rate also determines quanta removal through outdoor air exchange, in addition to inactivation of \emph{Mtb} in airborne particles. (6)~Quanta generation, diffusion, and removal determine the spatiotemporal quanta concentration, from which the personal risk of infection is calculated using the Wells-Riley equation.}
    \label{fig:modeling-flow}
\end{figure}

@@

(1)~We matched clinical and patient movement data to identify the spatiotemporal location of infectious TB patients within the clinic. We recorded the timestamp when the patient movement was detected in the registration area and matched this with the registration time in the clinical records. We excluded patient movements of less than five minutes and with less than five seconds in the registration area from the matching. Data were matched if there was a patient movement in the registration area followed by a entry in the clinical database, considering a maximum delay of 15 minutes. All diagnosed TB patients could be matched with patient movements this way. (2)~In addition to diagnosed TB patients, we also consider a number of undiagnosed (subclinical) TB patients. Prior studies suggest that the number of undiagnosed TB patients may be proportional to the number of diagnosed TB patients \cite{Berhanu2023CID,Moyo2022LancetID,Patterson2024PNAS}. Therefore, we model the number of undiagnosed TB patients with a multinomial distribution based on the counts of the daily number of diagnosed patients in the clinical data in October/November 2021 (\supp~Figure~\zref{fig:undiagnosed-distribution}): 6 days (12\%) with zero, 12 (41\%) with one, 8 (28\%) with two, 2 (7\%) with three, and 1 (3\%) with four TB patients visiting the clinic. Undiagnosed patients were randomly sampled among all clinic attendees with the same probability, excluding clinic staff who were not considered infectious. We further assumed that diagnosed and undiagnosed TB patients generated quantum at the same rate $q$, which was modeled using a lognormal prior distribution considering two different activity levels (\supp~Figure~\zref{fig:quanta-distribution}): median of 1.08\,quanta h$^{-1}$ (95\%-credible interval [CrI] 0.003\,quanta h$^{-1}$\,$-$\,386\,quanta h$^{-1}$) while sitting and 2.81\,quanta h$^{-1}$ (95\%-CrI 0.008\,quanta h$^{-1}$\,$-$\,987\,quanta h$^{-1}$) while walking\cite{Mikszewski2021GF,Buonanno2020EI,Banholzer2024PGPH}. Since mask wearing was mandatory for all clinic staff and attendees during the study, we assumed a mean reduction in $q$ by 75\% (95\%-CrI 56\%\,$-$\,86\%)\cite{Dharmadhikari2012AJRCCM,McCreesh2021BMJGlobalHealth}.

@@

(3)~We calculated room occupancy from the video sensor data and combined it with the CO$_2$ measurements to estimate air change rates by daytime using a transient mass balance model\cite{Batterman2017IJERPH}. (4)~The air change rate was related to the diffusion of quanta using, as an approximation, the empirical relationship between the eddy diffusion coefficient and the air change rate in CO$_2$ particles\cite{Cheng2011EnvSciTech,Foat2020BE}. Since airflow was not monitored in our study, we assumed that the quanta diffused radially from where it had been generated (\supp~Figure~\zref{fig:toy-example}).

@@

(5)~Quanta removal occurs through replacement of contaminated indoor air with fresh outdoor air, inactivation of bacteria in airborne particles, and gravitational settling. The removal rate is thus the sum of the air change rate, the inactivation rate of \emph{Mtb}, and the gravitational settling rate. The bacterial inactivation rate of \emph{Mtb} was modeled using a lognormal prior distribution (\supp~Figure~\zref{fig:undiagnosed-distribution}) with a median of 1\,quanta h$^{-1}$ (95\%-CrI 0.1\,quanta h$^{-1}$\,$-$\,7.1\,quanta h$^{-1}$)\cite{Loudon1969AMRRD,Lever2000LettersAppliedMicrobio,Gannon2007ResVetSci,Klein2014IJMyco}. We set the gravitational settling rate to zero, thereby ignoring sedimentation because the particles carrying \emph{Mtb}, which are in the size range of $1-7\mu$m\cite{Fennelly2020Lancet}, are airborne immediately or within seconds\cite{Vuorinen2020SafSci}.

@@

\section{Results}

>>>>

\title{\bfseries\LARGE\singlespacing{A novel spatiotemporal model to estimate \emph{Mycobacterium tuberculosis} transmission risks in crowded indoor settings using environmental, clinical, and patient movement data}}
% author list
\author[1]{Nicolas Banholzer}
\author[2]{Keren Middelkoop}
\author[2]{Juane Leukes}
\author[3]{Ernest Weingartner}
\author[1]{Remo Schmutz}
\author[1]{Kathrin Zürcher}
\author[1,4,5]{Matthias Egger}
\author[2]{Robin Wood}
\author[1*]{Lukas Fenner}

\affil[1]{Institute of Social and Preventive Medicine, University of Bern, Bern, Switzerland}
\affil[2]{Desmond Tutu HIV Centre, Department of Medicine, University of Cape Town, Cape Town, South Africa}
\affil[3]{Institute for Sensors and Electronics, University of Applied Sciences and Arts Northwestern Switzerland, Windisch, Switzerland}
\affil[4]{Population Health Sciences, University of Bristol, Bristol, United Kingdom}
\affil[5]{Centre for Infectious Disease Epidemiology and Research, University of Cape Town, Cape Town, South Africa}

@@

\begin{information}

\noindent\noindent\textbf{Supplementary Information:} Includes supplementary text, tables and figures.

\noindent\textbf{Supplementary Video:} Animated sequence of the spatiotemporally varying quanta concentration as an infectious patient enters the clinic on the morning of October 25 and moves through the waiting room.

\end{information}

@@

@@

%TC:newcounter abst Words in abstract
%TC:envir abstract [] abst
\begin{abstract}\normalfont
The risk of Mycobacterium tuberculosis (\emph{Mtb}) transmission can be high in crowded clinics. We developed a spatiotemporal model of airborne \emph{Mtb} transmission, based on the Wells-Riley equation. We collected environmental, clinical and person-tracking data in a South African clinic during COVID-19, when masks were compulsory and ventilation increased. We matched person tracks with clinical records to identify the spatiotemporal location of infectious TB patients. We modeled the concentration of infectious doses (quanta) and estimated the individual risk of infection. Over five days, video sensors detected 1,438 clinic attendees. CO$_2$ levels were low (median 431\,ppm, IQR 406\,ppm$-$458\,ppm); the quanta concentration was higher in the morning than in the afternoon, and highest in the waiting room. The estimated risk of infection per clinic attendee was 0.06\% (95\%-credible interval [CrI] 0.001\%$-$0.40\%). It increased with the number of close contacts with infectious patients and the time spent in the clinic, and was 1.28-fold (95\%-CrI 1.19$-$1.36) higher in scenarios without mask use and 1.74-fold (95\%-CrI 0.47$-$5.34) with pre-pandemic ventilation rates. Spatiotemporal modeling can identify high-risk areas and evaluate the impact of infection control measures in clinics.

% \noindent\textbf{Background:} Tuberculosis (TB) caused by \emph{Mycobacterium tuberculosis (Mtb)} is strictly airborne and thus the risk of transmission is higher in crowded indoor environments, especially healthcare facilities. Existing approaches to modeling the risk of airborne transmission typically assume that the airspace is well mixed, not considering that proximity to infectious individuals may increase the risk of transmission. We aimed to develop a  spatiotemporal model of airborne \emph{Mtb} transmission using environmental, clinical, and patient movement data to estimate TB transmission during the COVID-19 pandemic in healthcare facilities and to evaluate the impact of infection control measures.

% \noindent\textbf{Methods:} We collected environmental (CO$_2$ levels), clinical (patient visits and TB disease status), and patient movement data (anonymous patient tracking from video sensors) in a primary care clinic in Cape Town, South Africa, for five days in October/November 2021 during the COVID-19 pandemic when universal use of masks was compulsory and ventilation was increased. We matched patient movements with clinical records to identify the spatiotemporal location of  infectious TB patients. We modeled the spatiotemporal concentration of infectious doses (quanta) and estimated the patient-specific risk of infection using the Wells-Riley equation.

% \noindent\textbf{Results:} Clinical records registered 894 patients, four of whom were diagnosed with TB. Video sensors detected 1,438 clinic attendees with a median stay of 25\,min (interquartile range [IQR] 13\,min$-$46\,min). CO$_2$ levels were low (median 431\,ppm, IQR 406\,ppm$-$458\,ppm), compatible with high ventilation rates. The modeled risk of infection per clinic attendee was 0.059\% (95\%-credible interval [CrI] 0.001\%$-$0.402\%). The risk would have increased 1.28-fold (95\%-CrI 1.19$-$1.36) without mask use and 1.74-fold (95\%-CrI 0.47$-$5.34) with lower, pre-pandemic ventilation rates. Furthermore, the individual risk was significantly associated with the number of close-contact encounters and time spent in the clinic.

% \noindent\textbf{Conclusions:} Our novel spatiotemporal transmission model estimated the risk of \emph{Mtb} transmission in a South African clinic as low during the COVID-19 pandemic, mainly due to infection control measures. Our spatiotemporal modeling could be used to identify high-risk areas and evaluate the impact of infection control measures in clinics.

@@

\noindent\textbf{Keywords:} Mycobacterium tuberculosis, airborne infection, person tracking, spatiotemporal model
% maximum of 3-5 keywords
\newpage

@@

%TC:break main
\section*{Introduction}

% TB and the route of transmission
Tuberculosis (TB) remains one of the leading causes of death globally, with the South African region bearing a disproportionate share of the TB burden.  After two years of COVID-19-related disruption of TB prevention and control measures, the negative impact of the COVID-19 pandemic has been receding, as documented in the World Health Organization (WHO) 2023 global TB report\cite{WHO2023TBReport}. However, progress was insufficient to meet global TB targets\cite{WHO2023TBReport}. For example, investment in TB research was only about half of the USD 2~billion target\cite{WHO2023TBReport}.

\emph{Mycobacterium tuberculosis} (\emph{Mtb}), the causative agent of TB, is transmitted via respiratory particles in the exhaled air of infectious persons\cite{Rieder1999,Patterson2021Tuberculosis}. \emph{Mtb} is primarily carried in small particles of $1-7\mu$m\cite{Fennelly2020Lancet}, which can survive in the air for multiple hours\cite{Loudon1969AMRRD}. Survival of airborne pathogens is greater in crowded, poorly ventilated indoor environments\cite{Rieder1999,CPS2013Book,Nardell1991ARRD,Wang2021Science,Morawska2021}. Healthcare facilities are high-risk setting because infectious individuals are more likely to visit a clinic\cite{McCreesh2020IJTLD}. A modeling study in South Africa estimated that 4\% to 14\% of adult TB cases originate from primary care clinics\cite{McCreesh2022BMJGlobalHealth}.

% the Wells-Riley model and its limitations
The Wells-Riley model\cite{Riley1961Book} is widely used to estimate the risk of airborne transmission in a variety of indoor settings\cite{Haddrell2024NatCommun,Andrews2014JID,Taylor2016IJTLD,Hella2017JInfect,Zemouri2020JDR}, including primary care clinics\cite{Zurcher2022JID,McCreesh2021BMJGlobalHealth}. The model estimates the risk of infection based on the number of infectious individuals in space, the generation rate of infectious quanta (doses of pathogen-carrying particles), the breathing rate per person, and the outdoor air supply rate, which is often estimated from indoor CO$_2$ levels\cite{Rudnick2003IndoorAir,Richardson2014PONE,Haddrell2024NatCommun}. The Wells-Riley model and its variations\cite{Rudnick2003IndoorAir} assume a well-mixed airspace, meaning that the quanta concentration is the same throughout the room. However, this concentration is typically higher near the infectious source\cite{Wang2021Science,Vuorinen2020SafSci,Chen2020BuildEnv}, and previous studies have shown that the risk of TB infection is associated with proximity to these individuals\cite{Ko2004RiskAnal,Kenyon1996NEJM}.

% what this study adds
We developed a spatiotemporal extension of the Wells-Riley model to model the concentration of infectious quanta in space and time. We combined clinical and person-tracking data to determine the spatiotemporal location of TB patients and susceptible attendees at a primary care clinic in South Africa for five days in October/November 2021. We measured indoor CO$_2$ levels to model the spatial diffusion and removal of infectious quanta over time and estimated the risk of infection per clinic attendee. Finally, we assessed the impact of infection control measures during the study in response to the COVID-19 pandemic.

@@

\section*{Methods}

\subsection*{Study design and setting}

Based on a pilot study from 2019 in the same clinic\cite{Zurcher2022JID,Zurcher2020BMJ}, we collected environmental data (indoor CO$_2$ levels), clinical data (TB status), and video sensor data (person tracks) for five days during the COVID-19 pandemic in October/November 2021 (October 13, 15, 25, and November 4, 5) at a primary care clinic in Cape Town, South Africa. The clinic offers both TB and HIV services and other basic clinical services, Monday through Friday, from 8\,am to 4\,pm. It is located in a large settlement of formal and semi-formal housing where TB and HIV are highly prevalent\cite{Wood2007AMJRCCD,Middelkoop2011JAIDS}. We defined three areas within the clinic (\Cref{fig:floor-plan}): the waiting room (10.55m $\times$ 5.50m $\times$ 3.00m), corridor (12.45m $\times$ 2.20m $\times$ 2.50m), and TB room (4.75m $\times$ 3.50m $\times$ 3.00m). The TB room was mainly used for the treatment of diagnosed TB patients but also for the screening of suspected patients with respiratory symptoms who were ultimately not diagnosed with TB.

\begin{figure}[!htpb]
    \centering
    \includegraphics{doc/clinic-schematic-annotated-view.pdf}
    \caption{\textbf{Schematic view of the clinic}. Patients typically enter the clinic through the entrance on the bottom left, register at the reception, then wait in the waiting room or along the corridor until they see a doctor in one of the care rooms or the TB room. They exit the clinic through one of the exits in the waiting room or at the end of the corridor. Colored icons show the placement of the video sensors to track persons and the devices monitoring CO$_2$ levels.}
    \label{fig:floor-plan}
\end{figure}

\subsection*{Data collection}

We collected environmental and clinical data and data on the movements of patients in the clinic. Four devices monitored indoor CO$_2$ levels (CO$_2$ Logger CL11, Rotronic, Bassersdorf, Switzerland) in the waiting room, corridor, and TB room (\Cref{fig:floor-plan}), in parts per million (ppm) at 1-min intervals. On October~13, the corridor's CO$_2$ levels were missing due to a malfunctioning device and the waiting room's CO$_2$ levels were used for the corridor because their CO$_2$ measurements were similar on the other days.    
We extracted clinical data from the electronic patient registry for all patients who attended the clinic during the study. Clinical records included the date and time of arrival, TB diagnostic results, and date of TB treatment initiation. We defined infectious patients as individuals who tested bacteriologically positive for TB during the study or who had received TB treatment for less than four weeks. Patients on treatment for more than four weeks or with recently completed treatment were considered neither infectious nor susceptible. We also defined suspected TB patients as individuals who were tested for TB following symptom screening, but who were not diagnosed with TB.

We used an anonymous personal tracking system using video sensors (Xovis, Zollikofen, Switzerland) to monitor the movements of people (clinic staff, patients, and other visitors) throughout the clinic at 1-second intervals (\Cref{fig:floor-plan}). The resulting time-stamped tracking data consisted of a person’s height, their position recorded as x-y coordinates, and a unique ID for each person's track while in the clinic. People in the clinic could contribute multiple records if they moved out of the range of a sensor or were not recognized as a person. To address this, we created an R Shiny tool (\supp~Fig.~\zref{fig:shiny-app}) to rejoin interrupted tracks as described in \supp~Section~\zref{sec:setting-and-data}. We also labeled the tracks to identify the movements of clinic staff. \supp~Fig.~\zref{fig:tracking-examples} shows some sample tracks.

\subsection*{Spatiotemporal modeling approach}

Our spatiotemporal model is based on the Wells-Riley equation\cite{Riley1961Book}, which estimates the risk (probability in \%) of infection $P$ as
\begin{align}
    P = \frac{C}{S} = 1 - \exp\left(\frac{Ipqt}{Q}\right),
\end{align}
where $C$ is the number of diseased cases, $S$ is the number of susceptible cases, $I$ is the number of infectious people in the indoor space, $p$ is the breathing rate (m$^3$ h$^{-1}$), $q$ is the quantum (infectious dose) generation rate (quanta h$^{-1}$), $t$ is the exposure time (h), and $Q$ is the ventilation rate (m$^3$ h$^{-1}$). The probability of infection is estimated with a Poisson relation, considering the stochastic behavior of airborne infection, where one quantum corresponds to a 67\% risk of infection. The main assumption of the Wells-Riley model is a well-mixed airspace, which means that the quanta concentration $N = Iq/Q$ (quanta m$^{-3}$) is the same everywhere in the room.

We extended the Wells-Riley model, allowing spatial variation in the quanta concentration by combining patient-tracking data with clinical records. This way, the spatiotemporal location of infectious patients who generate quanta could be determined. A detailed description of our extension is provided in \supp~Section~\zref{sec:spattemp-model}, together with an illustrative example using simulated data (\supp~Fig.~\zref{fig:toy-example}). In the following, we describe the six main components of our modeling approach (\Cref{fig:modeling-flow}).

\begin{figure}[!htpb]
    \centering
    \includegraphics{doc/model-flow-chart.pdf}
    \caption{\textbf{Visual summary of the spatiotemporal modeling approach}. (1)~Clinical records and person tracks were linked based on their timestamps at the registration to determine the spatiotemporal location of diagnosed (infectious) TB patients. (2)~Both diagnosed and randomly selected undiagnosed TB patients among clinic attendees generate infectious quanta. (3)~Person-tracking and environmental data were combined to estimate the air change rate. (4)~Air change rate was related to the speed at which quanta diffuses indoors. (5)~The air change rate also determines quanta removal through outdoor air exchange, in addition to inactivation of \emph{Mtb} in airborne particles. (6)~Quanta generation, diffusion, and removal determine the spatiotemporal quanta concentration, from which the personal risk of infection is calculated using the Wells-Riley equation.}
    \label{fig:modeling-flow}
\end{figure}

@@

(1)~We matched clinical and person-tracking data to identify the spatiotemporal location of infectious TB patients within the clinic. We recorded the timestamp when a person was detected in the registration area and matched this with the registration time in the clinical records. We excluded tracks with a total duration of less than five minutes in the clinic and a duration of less than five seconds in the registration area from the matching. We allowed for a maximum delay of 15\,min between the track in the registration area and the entry in the clinical database. All diagnosed TB patients could be matched with person tracks this way.

(2)~In addition to diagnosed TB patients, we also assumed a number of undiagnosed (subclinical) TB patients. The number of undiagnosed TB patients may be proportional to the number of diagnosed TB patients \cite{Berhanu2023CID,Moyo2022LancetID,Patterson2024PNAS}. Therefore, we modeled the number of undiagnosed TB patients with a multinomial distribution based on the counts of the number of diagnosed patients in the clinical data in October/November 2021 (\supp~Fig.~\zref{fig:undiagnosed-distribution}): 6 days (12\%) with zero, 12 (41\%) with one, 8 (28\%) with two, 2 (7\%) with three, and 1 (3\%) with four diagnosed TB patients. Undiagnosed patients were randomly sampled among all clinic attendees with the same probability, excluding clinic staff who were not considered infectious. We further assumed that diagnosed and undiagnosed TB patients generated quantum at the same rate $q$, modeled using a lognormal prior distribution considering two different activity levels (\supp~Fig.~\zref{fig:quanta-distribution}): median of 1.08\,quanta h$^{-1}$ (95\%-credible interval [CrI] 0.003\,quanta h$^{-1}$\,$-$\,386\,quanta h$^{-1}$) while sitting and 2.81\,quanta h$^{-1}$ (95\%-CrI 0.008\,quanta h$^{-1}$\,$-$\,987\,quanta h$^{-1}$) while walking\cite{Mikszewski2021GF,Buonanno2020EI,Banholzer2024PGPH}. Since masks were mandatory for all clinic staff and attendees during the study, we assumed a mean reduction in $q$ by 75\% (95\%-CrI 56\%\,$-$\,86\%)\cite{Dharmadhikari2012AJRCCM,McCreesh2021BMJGlobalHealth}.

@@

(3)~We calculated room occupancy from the video sensor data and combined it with the CO$_2$ measurements to estimate air change rate during the morning (8\,am to 12\,am) and afternoon (12\,am to 4\,pm) of each clinic day using a transient mass balance model\cite{Batterman2017IJERPH}.

(4)~The air change rate was related to the diffusion of quanta using, as an approximation, the empirical relationship between the eddy diffusion coefficient and the air change rate in CO$_2$ particles\cite{Cheng2011EnvSciTech,Foat2020BE}. Since airflow was not monitored in our study, we assumed that the quanta diffused radially from where it had been generated.

@@

(5)~Quanta removal occurs through the replacement of contaminated indoor air with outdoor air, inactivation of bacteria in airborne particles, and gravitational settling. The removal rate is thus the sum of the air change rate, the inactivation rate of \emph{Mtb}, and the gravitational settling rate. We modeled the bacterial inactivation rate of \emph{Mtb} using a lognormal prior distribution (\supp~Fig.~\zref{fig:undiagnosed-distribution}) with a median of 1\,quanta h$^{-1}$ (95\%-CrI 0.1\,quanta h$^{-1}$\,$-$\,7.1\,quanta h$^{-1}$)\cite{Loudon1969AMRRD,Lever2000LettersAppliedMicrobio,Gannon2007ResVetSci,Klein2014IJMyco}. We set the gravitational settling rate to zero, ignoring sedimentation because the particles carrying \emph{Mtb}, which are in the size range of $1-7\mu$m\cite{Fennelly2020Lancet}, are airborne immediately or within seconds\cite{Vuorinen2020SafSci}.

@@

\section*{Results}

During one week (five open clinic days), 894 patients were registered in the clinical database: four patients were bacteriologically diagnosed with TB and 16 patients were suspected of having TB after symptom screening. Overall, 1,563 person tracks were detected based on the video sensor data; 125 likely from clinic staff (movements from and to staff rooms) and 1,438 from clinic attendees (registered patients and other visitors). Most movements were detected in the waiting room (\Cref{fig:input-data-descriptives}a) and during the morning (\Cref{fig:input-data-descriptives}b). Attendees spent about half an hour in the clinic (median 25\,min, interquartile range (IQR) 13\,min$-$46\,min), mostly in the waiting room (\Cref{fig:input-data-descriptives}c). Close-contact encounters were frequent (median 5, IQR 2$-$9 per attendee) and attendees were in close contact with at least one other person for more than half of the duration of their visit (median 68\%, IQR 41\%$-$87\%). The CO$_2$ levels in the clinic were low (median 436\,ppm, IQR 410\,ppm$-$462\,ppm (\supp~Fig.~\zref{fig:co2-levels}) and the corresponding air change rates were high throughout the study (\Cref{fig:input-data-descriptives}g), with a \eg median of 17 air changes per hour (IQR 9\,h$^{-1}$$-$23\,h$^{-1}$) in the densely occupied waiting room. Median temperature in the clinic was 22$^{\circ}$C (IQR 21$^{\circ}$C$-$23$^{\circ}$C) and relative humidity was 55\% (IQR 47\%$-$61\%).

\begin{figure}
    \centering
     \includegraphics{results/data/no-people-spatial.png}
    \includegraphics{results/data/no-people-over-time.png}
    \includegraphics{results/data/time-in-clinic.png}
    \includegraphics{results/data/air-changes-per-hour.png}
    \caption{\textbf{Person tracks and environmental conditions by clinic area.} Across clinic days and for the clinic waiting room, corridor and TB room: \textbf{(a)}~total number of tracks recorded per unit space, \textbf{(b)}~number of people in the clinic over time, \textbf{(c)}~time per clinic visit, and \textbf{(d)}~air change rates. Lines and ribbons show the mean plus/minus standard deviation. Boxplots show the medians as lines, interquartile ranges as boxes, ranges as whiskers, and outliers as dots.}
    \label{fig:input-data-descriptives}
\end{figure}
